\newcommand\mmskip{-6mm}

\newcommand\pubitem[6]{\item \href{#5}{#1} #2 \textit{#3} #4 preprint link: \href{#5}{#6}}

% 1. title
% 2. coauthor(s)
% 3. journal name
% 4. vol. num. pages. year.
% 5. url
% 6. short url
%_______________________________________________________________________________
    % Journal Articles
    {\it Journal Articles}\\[-6pt]
    \begin{enumerate}
      \pubitem
      {Universal algebraic methods for constraint satisfaction problems,}
      {with Clifford Bergman; to appear in}
      {Logical Methods in Computer Science {\small (LMCS)};}
      {}
      {https://arxiv.org/abs/1611.02867}
      {arXiv [cs.LO] 1611.02867}
        
      \pubitem
      {Polynomial-time tests for difference terms in idempotent varieties,}
      {with Freese and Valeriote; to appear in}
      {International Journal of Algebra \& Computation {\small (IJAC)};}
      {}
      {https://github.com/UniversalAlgebra/term-conditions/blob/master/ijac/diffTerm-ijac-r1-draft-20180905.pdf}
      {diffTerm-ijac-r1-draft-20180905.pdf}
      
      \pubitem 
      {Isotopic algebras with nonisomorphic congruence lattices}
      {(sole author)}  % no coauthor
      {Algebra Universalis}
      {\textbf{72}:295--298, 2014;} 
      {https://github.com/williamdemeo/Isotopy/blob/master/Isotopy-AU-2014.pdf}
      {Isotopy-AU-2014.pdf}
      
      \pubitem 
      {Expansions of finite algebras and their congruence lattices}
      {(sole author)}
      {Algebra Universalis}
      {\textbf{69}:257--278, 2013;}
      {https://github.com/williamdemeo/Overalgebras/blob/master/DeMeo-Expansions-AU-2013.pdf}
      {DeMeo-Expansions-AU-2013.pdf}

\end{enumerate}


    %_______________________________________________________________________________
    % Conference Proceedings
    {\it Refereed Conference Proceedings}\\[-6pt]
    \begin{enumerate}
      \setcounter{enumi}{4}
    %% \item \href{https://github.com/UniversalAlgebra/diff-term-existence/tree/master/LICS}%
    %%        {On deciding existence of difference terms},
    %%   submitted to \emph{LICS 2017}, Reykjavik, Iceland.

     \item \href{http://www.lulu.com/shop/kira-adaricheva-and-william-demeo-and-jennifer-hyndman/algebras-and-lattices-in-hawaii-honoring-ralph-freese-bill-lampe-and-jb-nation/paperback/product-23634583.html}{\it Proceedings of Algebras and Lattices in Hawaii 2018}, editor with K.~Adaricheva, J.~Hyndman.

    \item \href{https://github.com/williamdemeo/ISMA2004/raw/master/DeMeo-ISMA2004-FinalPaper.pdf}%
           {Topics in nonabelian harmonic analysis and \textsc{dsp}  applications},
      {\it Proceedings of the International Symposium on Musical Acoustics}, Nara, Japan 2004
      (best paper award).

    \item \href{https://github.com/williamdemeo/ICMC2002/raw/master/DeMeo-ICMC2002.pdf}%
           {Characterizing musical signals with Wigner-Ville interferences},
      {\it Proceedings of the International Computer Music Conference}, G\"{o}teborg, Sweden 2002.

     \item \href{https://williamdemeo.github.io/MSThesis}%
           {Approximating eigenvalues of large stochastic matrices},
      {\it Proceedings of the 8th Copper Mt.~Conference on Iterative Methods},
      Colorado, USA 1998.
    \end{enumerate}

\ifthenelse{\boolean{pagenumbers}}{
  \begin{center}
    \vskip-8mm
    \hfill (page 1 of 5)
  \end{center}
}{}
\newpage

~\vspace{-14mm}


    % In Preparation
    {\it Papers in Progress}\\[-8pt]
    \begin{itemize}
          \item[] \href{https://github.com/UniversalAlgebra/fg-free-lat}
            {A new characterization of fiber products of lattices},
            with P.~Mayr and N.~Ruskuc.
%            {\small Draft available at} \href{https://github.com/UniversalAlgebra/fg-free-lat}{{\small github.com/UniversalAlgebra/fg-fin-lat}}

\vskip\mmskip

    \item[] \href{https://github.com/UniversalAlgebra/fin-lat-rep}
           {Representing finite lattices as congruence lattices of finite algebras},
           with R.~Freese and P.~Jipsen.
          %  {\small Draft available at} \href{https://github.com/UniversalAlgebra/fin-lat-rep}{{\small github.com/UniversalAlgebra/fin-lat-rep}}
\end{itemize}
% \vskip-2mm

    {\it Books in Progress}\\[-8pt]
    \begin{itemize}
      \item[] \textsl{Algebras, Categories, Types: with computer-aided proofs}, with Hyeyoung Shin.

\vskip\mmskip
     \item[] \textsl{A Concise Course in Category Theory}, with Charlotte Aten and Venanzio Capretta.

% \vskip\mmskip

%      \item[] \href{http://www.math.hawaii.edu/~williamdemeo/PSRCA.html}{\it Problems in Real and Complex Analysis}.

% \vskip\mmskip
%      \item[] \href{http://www.math.hawaii.edu/~williamdemeo/PSRCA.html}{\it Problems in Groups and Rings}.

   \end{itemize}



      %% \end{itemize}
    %% \end{enumerate}
    %    \begin{enumerate}[leftmargin=15pt]
    %% \item[7a.] 
    %%   ``On a problem of Palfy and Saxl'' (latest draft: \href{http://github.com/williamdemeo/PalfySaxlProblem}{github.com/PalfySaxlProblem}).
    %% \vspace{-3mm}
    %% ``Subgroup lattice intervals of height two,'' with Ryan McCullough (in preparation).



        %_________________________________________________________________________________
    % Tech. Rep.
    %% \section{\mysidestyle Technical Reports}
    %% ``Notes on the finite lattice representation problem,'' \nosmall {\it University of Hawaii.  2011.}

    %% \vspace{-3mm}
    %% ``A fast algorithm for the multi-layer phase accumulation model of atmospheric turbulence,''\\
    %% \Textron{2005}

    %% \vspace{-3mm}
    %% ``Cone parameter estimation from simulated range/doppler data,'' \nosmall with Jon Guerin, \\
    %% \Textron{2005}

    %% \vspace{-3mm}
    %% ``Group harmonic analysis of turbulence-degraded images,'' \nosmall with Paul Billings, \\
    %% \Textron{2004}

    %% \vspace{-3mm}
    %% ``Speckle imaging on finite groups,'' with Paul Billings \\

    %% \vspace{-3mm}
    %% ``A Poisson model for object reconstruction''\\
    %% \Textron{2004}
